An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-09-05 nips_2016.tex
update paper and figures

\author{
  Haoyu Chen \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{haoyuchen@berkeley.edu}
  \And
  Daniel Seita \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{seita@berkeley.edu}
  \And
  Xinlei Pan \\
  Department of Bioengineering \\
  University of California, Berkeley \\
  \texttt{xinleipan@berkeley.edu}
  \And 
  Biye Jiang \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{bjiang@berkeley.edu}
  \And
  John Canny \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{canny@berkeley.edu}
}

@@

We conduct three sets of experiments to explore the benefits of our minibatch MH test and to
benchmark it with previous work. In Section~\ref{ssec:gaussians}, we show that our test enables
samples to converge to the posterior distribution of a heated Gaussian mixture model. In
Section~\ref{ssec:logistic}, we analyze its efficiency on logistic regression. In
Section~\ref{ssec:nets}, we apply the test for deep learning.
Appendices~\ref{app:gaussian},~\ref{app:logistic}, and~\ref{app:nnet} contain more detailed
information on these respective experiments.

@@

We start with a simple Gaussian mixture model, borrowing an experiment from~\cite{langevin_2011}.
The parameter is 2-D, $\theta = (\theta_1,\theta_2)$, and the parameter/data generation process is
\begin{equation}\label{eq:data_generation}
(\theta_1,\theta_2) \sim \mathcal{N}((0,0), {\rm diag}(\sigma_1^2, \sigma_2^2));\quad \quad x_i \sim
0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
\end{equation}
We set $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$. Fixing $\theta = (0,1)$, we draw 10000
data points so that the target distribution is $p(\theta)\prod_{i=1}^{10000}p(x_i\mid \theta)$, with
the prior based on the $\theta$ generation process in Equation~\ref{eq:data_generation}. This
results in rather sharp posterior modes and high ${\rm std}(\Delta')$ estimates, so we apply a
temperature $T=110$ to reduce ${\rm std}(\Delta')$. Taking logs, we get the target as shown in the
far left of Figure~\ref{fig:gauss_mix_1}.

We run MCMC with our MH test using minibatch size $m=200$. We also run this using standard
full-batch MCMC (``standard sampling'') with the standard test from Equation~\ref{eq:traditional}, and the method
from~\cite{cutting_mh_2014} (``adaptive sampling''). For the latter, we also use $m=200$ and
increment minibatches by that amount within an iteration if necessary. The tolerance for making a
decision is $\epsilon=0.1$. To make comparisons easier, all three use the same random walk proposer
with covariance $\Sigma={\rm diag}(0.03, 0.03)$. This is a poor proposer, but it is sufficient for
our purposes as the quality of the samples will be primarily due to the MH test. All methods are run
10000 times to collect 10000 samples.

\begin{figure}[t]
    \centering
    \includegraphics[width=1\linewidth]{cloud_v01.png}
    \caption{
    The log posterior contours (temperature 110) and three scatter plots of sampled $\theta$ values.
    }
    \label{fig:gauss_mix_1}
    %\vspace{-10pt}
\end{figure}

\begin{figure}[t]
    \centering
    \includegraphics[width=1\linewidth]{adaptive_and_ours_information_v01.png}
    \caption{
    Histograms of minibatch sizes and ${\rm std}(\Delta')$ values, along with ``${\rm std}(\Delta')$
    violations'' (right).
    }
    \label{fig:gauss_mix_2}
    %\vspace{-10pt}
\end{figure}

\begin{figure}[t]
    \centering
    \includegraphics[width=1\linewidth]{CI_v01.png}
    \caption{
    Histograms of ${\rm std}(\Delta')$, confidence interval of ${\rm std}(\Delta')$ (middle), along with $\rho / \sigma^3$ (right).
    }
    \label{fig:gauss_mix_3}
    %\vspace{-10pt}
\end{figure}

Figure~\ref{fig:gauss_mix_1} shows scatter plots of the resulting $\theta$ samples for the four
methods, with darker regions indicating a greater density of points. The first three methods obtain the
same rough form of the posterior, so our MH test can indeed result in the same posterior as the
other two methods. However, the fourth method (i.e. just treat the $\Delta'$ which comes from minibatch as $\Delta$ which comes from the complete data) reveals wider spread distribution than the posterior. From Figure~\ref{fig:gauss_mix_1} we illustrate our proposed method can obtain the samples which have almost the same distribution as the standard MH test from the complete data, adaptive minibatch test, and the true posterior. Meanwhile, we show the benefits of our minibatch method compared with the naive minibatch method.

Figure~\ref{fig:gauss_mix_2} (to the left) shows a histogram of the final minibatch sizes used by
the adaptive subsampling method each iteration (on a log scale for readability).  Usually, the method
can make a decision with 200 samples. Occasionally, however, it must use all 10000 points, resulting
in lots of wasted computation.  For this particular experiment, that happened eight times.
The third plot in Figure~\ref{fig:gauss_mix_2} investigates the number of times ${\rm std}(\Delta')
\ge 1.2$ for only consuming one minibatch data. For six minibatch sizes (50, 100, 150, 200, 250, and 300), we ran MCMC with
our MH test five times and averaged the number of ``violations.'' As the size increases, the
number of violations decreases, as expected.

Figure~\ref{fig:gauss_mix_3} (to the left) shows a histogram of ${\rm
  std}(\Delta')$. Figure~\ref{fig:gauss_mix_3} (to the middle) shows
the confidence interval of our estimated ${\rm std}(\Delta')$ at each
iteration \footnote{In order to show the confidence interval clearly,
  we use moving smooth to smooth the confidence interval, and plot
  each value by interval of 50 iterations}. The blue line represents
the standard deviation comes from the complete data, which can be
treated as the ``true'' ${\rm std}(\Delta')$, the green line
represents our estimation of ${\rm std}(\Delta')$ from the minibatch,
and the two red dash lines represent the confidence interval of ${\rm
  std}(\Delta')$ by adding / subtracting one ${\rm
  std(std(}\Delta'))$. This figure indicates that our estimation of
${\rm std}(\Delta')$ from one minibatch data is
accurate. Figure~\ref{fig:gauss_mix_3} (to the right) shows the value
of $\rho/\sigma^3$ at each iteration.

@@

We next use logistic regression for the binary classification of 1s versus 7s in the MNIST
dataset~\cite{lecun-mnisthandwrittendigit-2010}. The data has 12007 and 1000 training and testing
points, respectively (we used a random subset of the test data). The proposer is again a random walk
with covariance matrix $0.1I$ for the $784\times 784$ identity matrix $I$. We set the posterior
temperature at $T=3000$. We set the minibatch size $m=50$ and compare with adaptive sampling with
tolerance 0.02.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=1\linewidth]{exp2.png}
    \caption{Logistic regression performance (accuracy/log likelihood) and minibatch size analysis.}
    \label{fig:logistic_fig}
    \vspace{-10pt}
\end{figure}

\begin{figure}[t]
    \centering
    \includegraphics[width=1\linewidth]{CI_v02.png}
    \caption{
    Histograms of ${\rm std}(\Delta')$, confidence interval of ${\rm std}(\Delta')$ (middle), along with $\rho / \sigma^3$ (right).
    }
    \label{fig:exp_3}
    %\vspace{-10pt}
\end{figure}

The third plot of Figure~\ref{fig:logistic_fig} shows performance based on minibatch size. The
parameter is evaluated on the test set after training iterations 700, 1000, and 1500. For those
three cases, when the minibatch size increases beyond about 10, the performance of the sampled data
will not improve with a large minibatch size. This makes sense because the error of our MH test is
proportional to our estimation error of ${\rm std}(\Delta')$, and as the minibatch increases, ${\rm
std}(\Delta')$ decreases. Beyond a certain standard deviation (or minibatch) threshold, the error in
our minibatch MH test is negligible.

Similar as the Subsection~\ref{ssec:gaussians}, we use Figure~\ref{fig:exp_3} to reveal the accuracy of our estimation of ${\rm std}(\Delta')$ and stability of $\rho/\sigma^3$.

\subsection{Neural Network Optimization}\label{ssec:nets}

@@

>>>>

\author{
  John Canny \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{canny@berkeley.edu}
  \And
  Haoyu Chen \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{haoyuchen@berkeley.edu}
  \And
  Daniel Seita \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{seita@berkeley.edu}
  \And
  Xinlei Pan \\
  Department of Bioengineering \\
  University of California, Berkeley \\
  \texttt{xinleipan@berkeley.edu}
  \And 
  Biye Jiang \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{bjiang@berkeley.edu}
 }

@@

We conduct two sets of experiments to explore the benefits of our minibatch MH test and to
benchmark it with previous work. In Section~\ref{ssec:gaussians}, we show that our test enables
samples to converge to the posterior distribution of a heated Gaussian mixture model. In
Section~\ref{ssec:logistic}, we analyze its efficiency on logistic regression. 
%In %Section~\ref{ssec:nets}, we apply the test for deep learning. 
Appendices~\ref{app:gaussian},~\ref{app:logistic} %and~\ref{app:nnet} 
contain more detailed information on these respective experiments.

@@

We start with a simple Gaussian mixture model, borrowing an experiment from~\cite{langevin_2011}.
The parameter is 2-D, $\theta = (\theta_1,\theta_2)$, and the parameter/data generation process is
\begin{equation}\label{eq:data_generation}
(\theta_1,\theta_2) \sim \mathcal{N}((0,0), {\rm diag}(\sigma_1^2, \sigma_2^2));\quad \quad x_i \sim
0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
\end{equation}
We set $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$. Fixing $\theta = (0,1)$, we draw 1,000,000
data points so that the target distribution is $p(\theta)\prod_{i=1}^{1,000,000}p(x_i\mid \theta)$, with
the prior based on the $\theta$ generation process in Equation~\ref{eq:data_generation}. This
results in rather sharp posterior modes and high ${\rm std}(\Delta')$ estimates, so we apply a
temperature $T=10,000$ to reduce ${\rm std}(\Delta')$. Taking logs, we get the target as shown in the
far left of Figure~\ref{fig:gauss_mix_1}.

We run MCMC with our MH test using minibatch size $m=100$. We also run this using standard
full-batch MCMC (``standard MCMC'') with the standard test from Equation~\ref{eq:traditional}, the method
from~\cite{cutting_mh_2014} (``Cut MH'')  and the method from ~\cite{icml2014c1_bardenet14}. For the method in \cite{cutting_mh_2014}, we also use $m=100$ and
increment minibatches by that amount within an iteration if necessary. As for the method in ~\cite{icml2014c1_bardenet14}, we use $m=100$ and increase the minibatch size geometrically with a ratio of $\gamma = 1.5$. The tolerance for making a
decision in \cite{cutting_mh_2014} is $\epsilon=0.005$, and the error bound control parameter in ~\cite{icml2014c1_bardenet14} is $p = 2, \delta = 0.01$. To make comparisons easier, all three use the same random walk proposer
with covariance $\Sigma={\rm diag}(0.3, 0.3)$. This is a poor proposer, but it is sufficient for
our purposes as the quality of the samples will be primarily due to the MH test. All methods are run
5000 times to collect 5000 samples.

\begin{figure}[t]
    \centering
    % \includegraphics[width=1\linewidth]{cloud_v01.png}
    \includegraphics[width=1\linewidth]{posterior_of_gaussian.png}
    \caption{
    The log posterior contours and scatter plots of sampled $\theta$ values using different methods. 
    }
    \label{fig:gauss_mix_1}
    %\vspace{-10pt}
\end{figure}

\begin{figure}[t]
    \centering
   % \includegraphics[width=1\linewidth]{adaptive_and_ours_information_v01.png}
   \includegraphics[width=1\linewidth]{minibatch_size_gaussian.png}
    \caption{
    Histograms of minibatch sizes %and ${\rm std}(\Delta')$ values, along with ``${\rm std}(\Delta')$
   % violations'' (right).
    }
    \label{fig:gauss_mix_2}
    %\vspace{-10pt}
\end{figure}

\begin{figure}[t]
	\centering
	\includegraphics[width=\linewidth]{GaussianMixtureResult/first_second_part_gaussian.png}
	\caption{Value of first and second part of equation (9) in ~\cite{icml2014c1_bardenet14} with respect to actual minibatch size}
	\label{fig:gauss_bardenet}
\end{figure}

%\begin{figure}[t]
    %\centering
    %\includegraphics[width=1\linewidth]{CI_v01.png}
   % \caption{
   % Histograms of ${\rm std}(\Delta')$, confidence interval of ${\rm std}(\Delta')$ (middle), along with $\rho / \sigma^3$ (right).
    %}
   % \label{fig:gauss_mix_3}
    %\vspace{-10pt}
%\end{figure}

Figure~\ref{fig:gauss_mix_1} shows scatter plots of the resulting $\theta$ samples for the four
methods, with darker regions indicating a greater density of points. The first three methods obtain the
same rough form of the posterior, so our MH test can indeed result in the same posterior as the
other two methods. %However, the fourth method (i.e. just treat the $\Delta'$ which comes from minibatch as $\Delta$ which comes from the complete data) reveals wider spread distribution than the posterior. From Figure~\ref{fig:gauss_mix_1} we illustrate our proposed method can obtain the samples which have almost the same distribution as the standard MH test from the complete data, adaptive minibatch test, and the true posterior. Meanwhile, we show the benefits of our minibatch method compared with the naive minibatch method.  
Actually, based on a measure of divergence from the ground-truth distribution, our method is better than the other three methods in terms of this measurement. The measurement is to calculate the poisson likelihood of the MCMC-generated parameters over a parameter range, with $\theta_1 \in (-1.5, 2.5)$ and $\theta_2 \in (-2.5, 2.5)$. The log likelihood value for our method, standard MCMC, cut MH method and adaptive MH method are -2144.6, -2783.5, -4007.4, -2598.4, respectively.

Figure~\ref{fig:gauss_mix_2} shows a histogram of the final minibatch sizes used by
the four methods in each iteration. It is obvious that our methods consumes significantly less data, with most of minibatch sizes smaller than 1000, with a average minibatch size of 420. The other three methods occasionally consumes a large proportion of the entire data set, and the average minibatch sizes are 37785, 16378, 16755 for standard MCMC, cut MH method, adaptive MH method, respectively. In terms of average minibatch size, our method is nearly 90 times faster than standard MCMC method, and nearly 39 times faster the cut MH method, and nearly 40 times faster than the adaptive MH method.

It is useful to notice that, in ~\cite{icml2014c1_bardenet14}, they need to evaluate the likelihood term $C_{\theta,\theta'}$ over the entire the data set at every iteration in order to calculate the error bound in equation (9) in ~\cite{icml2014c1_bardenet14}. In Figure ~\ref{fig:gauss_bardenet}, the value of the first part and second part of equation (9) in ~\cite{icml2014c1_bardenet14} is shown with respect to the size of actual minibatch data used. The second part which contains  the $C_{\theta,\theta'}$ term is larger than the first part, which means that the second part is non-negligible and has to be calculated at every iteration. This is one of the very time-consuming parts in this method.

%Usually, the method can make a decision with 200 samples. Occasionally, however, it must use all 10000 points, resulting in lots of wasted computation.  For this particular experiment, that happened eight times.
%The third plot in Figure~\ref{fig:gauss_mix_2} investigates the number of times ${\rm std}(\Delta')
%\ge 1.2$ for only consuming one minibatch data. For six minibatch sizes (50, 100, 150, 200, 250, and 300), we ran MCMC with
% our MH test five times and averaged the number of ``violations.'' As the size increases, the
% number of violations decreases, as expected.

\iffalse
Figure~\ref{fig:gauss_mix_3} (to the left) shows a histogram of ${\rm
  std}(\Delta')$. Figure~\ref{fig:gauss_mix_3} (to the middle) shows
the confidence interval of our estimated ${\rm std}(\Delta')$ at each
iteration \footnote{In order to show the confidence interval clearly,
  we use moving smooth to smooth the confidence interval, and plot
  each value by interval of 50 iterations}. The blue line represents
the standard deviation comes from the complete data, which can be
treated as the ``true'' ${\rm std}(\Delta')$, the green line
represents our estimation of ${\rm std}(\Delta')$ from the minibatch,
and the two red dash lines represent the confidence interval of ${\rm
  std}(\Delta')$ by adding / subtracting one ${\rm
  std(std(}\Delta'))$. This figure indicates that our estimation of
${\rm std}(\Delta')$ from one minibatch data is
accurate. Figure~\ref{fig:gauss_mix_3} (to the right) shows the value
of $\rho/\sigma^3$ at each iteration. 
\fi

@@

We next use logistic regression for the binary classification of 1s versus 7s in the MNIST
dataset~\cite{lecun-mnisthandwrittendigit-2010}. The data has 12007 and 1000 training and testing
points, respectively (we used a random subset of the test data). The proposer is again a random walk
with covariance matrix $0.05I$ for the $784\times 784$ identity matrix $I$. We set the posterior
temperature at $T=1000$. We set the minibatch size $m=100$ and compare with ~\cite{cutting_mh_2014} with
tolerance 0.005 and with ~\cite{icml2014c1_bardenet14} with error bound control parameter $p = 2, \delta = 0.01$.

@@

\begin{figure}[t]
    \centering
    \begin{subfigure}{.5\textwidth}
     \centering
     \includegraphics[width=1\linewidth]{LogisticRegressionResult/acc_logistic}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=1\linewidth]{LogisticRegressionResult/ll_logistic}
\end{subfigure}
\caption{Logistic regression performance (accuracy/log likelihood) and minibatch size analysis}
\label{fig:logistic_fig}
\vspace{-10pt}
\end{figure}

\begin{figure}[t]
	\centering
	\includegraphics[width=1\linewidth]{LogisticRegressionResult/minibatch_size_logistic.png}
	\caption{Histograms of minibatch sizes of logistic regression parameter estimation}
	\label{fig:logistic_minibatch}
\end{figure}

%\begin{figure}[t]
  %  \centering
    %\includegraphics[width=1\linewidth]{CI_v02.png}
    %\caption{
   % Histograms of ${\rm std}(\Delta')$, confidence interval of ${\rm std}(\Delta')$ (middle), along with $\rho / \sigma^3$ (right).
    %}
    %\label{fig:exp_3}
    %\vspace{-10pt}
%\end{figure}

The plot of Figure ~\ref{fig:logistic_minibatch} shows the histogram of minibatch sizes. Choosing an initial minibatch size of 100, 
the standard MCMC, cut MH method, adaptive MH method achieves of final average minibatch size of 910, 585, and 2731 respectively, while our method achieves 
that of 100, showing significant better performance than the benchmark test methods.

%The third plot of Figure~\ref{fig:logistic_minibatch} shows performance based on minibatch size. The
%parameter is evaluated on the test set after training iterations 700, 1000, and 1500. For those
%three cases, when the minibatch size increases beyond about 10, the performance of the sampled data
%will not improve with a large minibatch size. This makes sense because the error of our MH test is
%proportional to our estimation error of ${\rm std}(\Delta')$, and as the minibatch increases, ${\rm
%std}(\Delta')$ decreases. Beyond a certain standard deviation (or minibatch) threshold, the error in
%our minibatch MH test is negligible.

%Similar as the Subsection~\ref{ssec:gaussians}, we use Figure~\ref{fig:exp_3} to reveal the accuracy of our estimation of ${\rm std}(\Delta')$ and stability of $\rho/\sigma^3$.

\iffalse
\subsection{Neural Network Optimization}\label{ssec:nets}

@@

\fi
